---
title: "My Second Species"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we run our setup function. 
```{r}
source("/home/tvnguy28/ColbyForecasting/setup.R")
```

Our second species, Ditylum brightwelli, is an autotrophic unicellular phytoplankton and is preyed on by our first species, the copepod Temora longicornis. To help us keep track of changes to the model input, let's make a configuration file for our second species. 

```{r}
SPECIES2 = "Ditylum brightwellii"
```

Let's load the prequisites required for our model. `coast` represents the coastline of the gulf of maine, which is helpful for visualization. `present` is the Brickman prediction for current temperature and scenarios. From the Brickman dataset, we also pull `mask` which defines the region of the Gulf of Maine for which we have a prediction. This is helpful for filtering out entries and records whose spatial and temporal coords are outside of Brickman's scope.  

```{r load_prequisites}
coast = read_coastline()
present = brickman_database() |>
  dplyr::filter(scenario == "PRESENT", interval == "mon") |> read_brickman()
mask = brickman_database() |>
  filter(scenario == "STATIC", var == "mask") |> read_brickman()
```

We made a custom function to read directly from our second species. Let's read our OBIS entries from here. 
```{r load obs}
obs = read_Dbrightwelli()
```

Perform thinning on our data. 
```{r thin_observations}
thinned_obs = sapply(month.abb,
               function(mon){ 
                 temp_x = obs |> filter(month == mon)
                 if(nrow(temp_x) == 0) return(NULL)
                 thin_by_cell(temp_x, mask)
               }, simplify = FALSE) |>
  dplyr::bind_rows() 
```

Quickly tally up our observations. Later on, we use this to determine the average count of observations, which we use for our background sampling rate. 
```{r all_observations_count}
LON0 = -67
LAT0 = 46
all_counts = count(st_drop_geometry(obs), month) # counting is faster without spatial baggage
all_counts
```
Calculate the background points per month. 
```{r n_background_per_month}
nback_avg = mean(all_counts$n) |>
  round()
nback_avg
```

For each month, generate random background points. 
```{r sample_background}
obsbkg <- sapply(month.abb,
  function(mon) {
    sample_background(
      thinned_obs |> filter(month == mon),
      mask,
      method = "random",  # <-- it needs to know it's a bias map
      return_pres = TRUE, # <-- give me the obs back, too
      n = nback_avg) |>   # <-- how many points
      mutate(month = mon, .before = 1)
  }, simplify = FALSE) |>
  bind_rows() |>
  mutate(month = factor(month, levels = month.abb))
obsbkg
```

## Covariates 

We now move on to prepare our covariates, first we read from the present Brickman data and choose to filter out collinear data. 

```{r filter_collinear}
keep = filter_collinear(present, method = "cor_caret", cutoff = 0.65)
keep
```

Add depth and month to list of covariates. 


```{r add_depth_and_month}
keep = c("depth", "month", keep)
```

Earlier we saved our model input, with background points in obsbkg. Let's make a copy and call it model_input
```{r present_model_input}
model_input = obsbkg
```

Add depth to list of extract Brickman variables
```{r}
present = read_brickman(add = c("depth"))
```

We extract the environmental covariates. 
```{r extract}
variables = extract_brickman(present, model_input, form = "wide")
variables
```

We don't need the .id column. 
```{r wrangle_variables}
variables = variables |>
  mutate(class = model_input$class) |>    # the $ extracts a column 
  select(-.id)                            # the minus means "deselect" or "drop"
variables
```

Let's look at a comparison of the covariates distribution between our presence and background points. 
```{r plot_pres_vs_bg, warning = FALSE}
plot_pres_vs_bg(variables |> select(-month), "class")
```
```{r write_cfg}
cfg = list(
  version = "v1",
  scientificname = SPECIES2,
  background = "average of observations per month",
  bias = "random",
  thinning = "true",
  keep_vars =  keep)


write_configuration(cfg)          
```

We wrote the configuration to disk, now it's time to read it back. 
```{r}
rcfg = read_configuration(SPECIES2, "v1")
```


And finally, write everything we computed to the disk. 
```{r write_model_input}
write_model_input(variables, scientificname = SPECIES2, version = "v1")

```

```{r read model input}
# model_input = read_model_input(scientificname = SPECIES2)
# example read model input
```
