---
title: "My Second Species"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we run our setup function. 
```{r}
source("/home/tvnguy28/ColbyForecasting/setup.R")
```

Our second species, Ditylum brightwelli, is an autotrophic unicellular phytoplankton and is preyed on by our first species, the copepod Temora longicornis 

```{r}
SPECIES2 = "Ditylum brightwellii"

cfg = list(
  version = "v1",
  scientificname = SPECIES2,
  background = "average of observations per month",
  bias = "random",
  thinning = "true",
  keep_vars =  keep)


write_configuration(cfg)            
```


```{r}
rcfg = read_configuration(SPECIES2, "v1")
```

```{r load_prequisites}
coast = read_coastline()
present = brickman_database() |>
  dplyr::filter(scenario == "PRESENT", interval == "mon") |> read_brickman()
mask = brickman_database() |>
  filter(scenario == "STATIC", var == "mask") |> read_brickman()
```

```{r load obs}
obs = read_Dbrightwelli()
```

```{r thin_observations}
thinned_obs = sapply(month.abb,
               function(mon){ 
                 temp_x = obs |> filter(month == mon)
                 if(nrow(temp_x) == 0) return(NULL)
                 thin_by_cell(temp_x, mask)
               }, simplify = FALSE) |>
  dplyr::bind_rows() 
```

```{r all_observations_count}
LON0 = -67
LAT0 = 46
all_counts = count(st_drop_geometry(obs), month) # counting is faster without spatial baggage
all_counts
```

```{r n_background_per_month}
nback_avg = mean(all_counts$n) |>
  round()
nback_avg
```

```{r sample_background}
obsbkg <- sapply(month.abb,
  function(mon) {
    sample_background(
      thinned_obs |> filter(month == mon),
      mask,
      method = "random",  # <-- it needs to know it's a bias map
      return_pres = TRUE, # <-- give me the obs back, too
      n = nback_avg) |>   # <-- how many points
      mutate(month = mon, .before = 1)
  }, simplify = FALSE) |>
  bind_rows() |>
  mutate(month = factor(month, levels = month.abb))
obsbkg
```

## Covariates 

We now move on to prepare our covariates, first we read from the present Brickman data and choose to filter out collinear data. 

```{r filter_collinear}
keep = filter_collinear(present, method = "cor_caret", cutoff = 0.65)
keep
```

Add depth and month to list of covariates. 


```{r add_depth_and_month}
keep = c("depth", "month", keep)
```

Earlier we saved our model input, with background points in obsbkg. Let's make a copy and call it model_input
```{r present_model_input}
model_input = obsbkg
```

We extract the environmental covariates. 
```{r extract}
variables = extract_brickman(present, model_input, form = "wide")
variables
```

We don't need the .id column. 
```{r wrangle_variables}
variables = variables |>
  mutate(class = model_input$class) |>    # the $ extracts a column 
  select(-.id)                            # the minus means "deselect" or "drop"
variables
```

Let's look at a comparison of the covariates distribution between our presence and background points. 
```{r plot_pres_vs_bg, warning = FALSE}
plot_pres_vs_bg(variables |> select(-month), "class")
```

```{r write_model_input}
write_model_input(variables, scientificname = SPECIES2, version = "v1")

```

```{r read model input}
# model_input = read_model_input(scientificname = SPECIES2)
# example read model input
```