---
title: "Models"
output: html_document
---


```{r setup}
source("setup.R")
```

```{r second_species_name}
SPECIES2 = "Ditylum brightwellii"
```

# Load data - choose a month and sampling approach

```{r load_data}
cfg = read_configuration(scientificname = SPECIES2, version = "v1")
model_input = read_model_input(scientificname = SPECIES2, 
                               version = "v1",
                               log_me = c("depth", "Xbtm")) |>
  dplyr::mutate(month = month_as_number(.data$month)) |>
  select(all_of(c("class", cfg$keep_vars)))
```

# Splitting the data into groups

## Initial split into training and testing groups

```{r initial_split}
model_input_split = spatial_initial_split(model_input, 
                        prop = 1 / 5,     # 20% for testing
                        strategy = spatial_block_cv) # see ?spatial_block_cv
model_input_split
```


```{r initial_split_plot}
autoplot(model_input_split)
```

## Split the training group into regional folds


```{r cv_training}
tr_data = training(model_input_split)
cv_tr_data <- spatial_block_cv(tr_data,
  v = 5,     
  cellsize = grid_cellsize(model_input),
  offset = grid_offset(model_input) + 0.00001
)
autoplot(cv_tr_data)
```


# Build a recipe

```{r recipe}
one_row_of_training_data = dplyr::slice(tr_data,1)
rec = recipe(one_row_of_training_data, formula = class ~ .)
rec
```

```{r recipe_summary}
summary(rec)
```

# Create a workflow


```{r make_workflow}
wflow = workflow_set(
  
  preproc = list(default = rec), # not much happening in our preprocessor
  
  models = list(                 # but we have 4 models to add
    
      # very simple - nothing to tune
      glm = logistic_reg(
          mode = "classification") |>
        set_engine("glm"),
      
      # two knobs to tune
      rf = rand_forest(
          mtry = tune(),
          trees = tune(),
          mode = "classification") |>
        set_engine("ranger", 
                   importance = "impurity"),
      
      # so many things to tune!
      btree = boost_tree(
          mtry = tune(), 
          trees = tune(), 
          tree_depth = tune(), 
          learn_rate = tune(), 
          loss_reduction = tune(), 
          stop_iter = tune(),
          mode = "classification") |>
        set_engine("xgboost"),
    
      # just two again
      maxent = maxent(
          feature_classes = tune(),
          regularization_multiplier = tune(),
          mode = "classification") |>
        set_engine("maxnet")
  )
)
wflow
```


## Chosing metrics for measuring how well the models do


```{r metrics}
metrics = sdm_metric_set(yardstick::accuracy)
metrics
```


# Fit the models to the various recipes

### Iterate to find various hyperparameter sets

```{r fit, warning = FALSE}
wflow <- wflow |>
  workflow_map("tune_grid",
    resamples = cv_tr_data, 
    grid = 3,
    metrics = metrics, 
    verbose = TRUE)
```
Here's a quick plot of the outputs, with each metric being showcased.
```{r plot_wflow}
autoplot(wflow)
```

### Choose the best set of hyperparameters for each model

```{r select_best}
model_fits = workflowset_selectomatic(wflow, model_input_split,
                                  filename = "Ditylum_brightwellii-v1-model_fits",
                                  path = data_path("models"))
model_fits
```

## Exploring the collection of model fit results


### A table of metrics

We can quickly pull together summary statistics.

```{r model_fit_metrics}
model_fit_metrics(model_fits)
```

### Confusion matrices and accuracy

```{r model_fit_confmat}
model_fit_confmat(model_fits)
```
### ROC/AUC


```{r model_fit_roc_auc}
model_fit_roc_auc(model_fits)
```

### Variable importance

```{r model_fit_vip}
model_fit_varimp_plot(model_fits)
```

## Exploring a single model fit result

Let's pull out one of the model fits and look at it's parts.

```{r random_forest}
rf = model_fits |>
  filter(wflow_id == "default_rf")
rf
```

#### `splits`

These are the allocations of training and testing elements in the original dataset. We can see it looks like the same distribution as our initial split (above).

```{r rf_splits}
autoplot(rf$splits[[1]])
```

#### `.metrics`


```{r rf_metrics}
rf$.metrics[[1]]
```

#### `.predictions`

Here we can see the **testing** data predictions. Ultimately we are interested in `.pred_class` and `class`, but there's a lot of data to explore in here.  

```{r rf_preds}
rf$.predictions[[1]]
```

#### `.workflow`

The workflow is the encapsulation of the model formula form the recipe, the tuned hyperparameters and well as any fitting coefficients for the mode.  When we print it we get a summary statement about the model.

```{r rf_workflow}
rf$.workflow[[1]]
```


### Partial dependence plot

Partial dependence reflects the relative contribution of each variable influence over it's full range of values.  The output is a grid grid of plots showing the relative distribution of the variable (bars) as well as the relative influence of the variable (line). 
```{r pd_plot_gtree}
model_fit_pdp(model_fits, wid = "default_btree", title = "Boosted Tree")
```

```{r pd_plot_rf}
model_fit_pdp(model_fits, wid = "default_rf", title = "Random Forest")
```

```{r pd_plot_maxent}
model_fit_pdp(model_fits, wid = "default_maxent", title = "Max Entropy")
```

```{r pd_plot_glm}
model_fit_pdp(model_fits, wid = "default_glm", title = "Generalized Linear Model")
```

