---
title: "Make All For T. longicornis"
output: github_document
---

First we run our setup function. 
```{r}
source("./setup.R")
```

```{r load_prequisites}
coast = read_coastline()
present = brickman_database() |>
  dplyr::filter(scenario == "PRESENT", interval == "mon") |> read_brickman()
mask = brickman_database() |>
  filter(scenario == "STATIC", var == "mask") |> read_brickman()
```

```{r load obs}
obs = read_Tlongicornis()
```

Perform thinning on our data. 
```{r thin_observations}
thinned_obs = sapply(month.abb,
               function(mon){ 
                 temp_x = obs |> filter(month == mon)
                 if(nrow(temp_x) == 0) return(NULL)
                 # thin_by_dist(temp_x, 20000)
                 temp_x
               }, simplify = FALSE) |>
  dplyr::bind_rows() 
```

Quickly tally up our observations. Later on, we use this to determine the average count of observations, which we use for our background sampling rate. 
```{r all_observations_count}
LON0 = -67
LAT0 = 46
all_counts = count(st_drop_geometry(obs), month) # counting is faster without spatial baggage
```
Calculate the background points per month. 
```{r n_background_per_month}
nback_avg = mean(all_counts$n) |>
  round()
nback_avg
```

For each month, generate random background points. 
```{r sample_background}
obsbkg <- sapply(month.abb,
  function(mon) {
    sample_background(
      thinned_obs |> filter(month == mon),
      mask,
      method = "random",  # <-- it needs to know it's a bias map
      return_pres = TRUE, # <-- give me the obs back, too
      n = nback_avg) |>   # <-- how many points
      mutate(month = mon, .before = 1)
  }, simplify = FALSE) |>
  bind_rows() |>
  mutate(month = factor(month, levels = month.abb))
```

View whether the trace line is still visible. 
```{r plot_presence_background}
ggplot() +
  geom_sf(data = obsbkg, 
          mapping = aes(),
          alpha =  0.4, shape = "circle small", size = 1) +
  geom_sf(data = coast, col = "orange")  + 
  labs(x = "Longitude", y = "Latitude", title = "All") +   
  theme_bw() +  # <- make a simple white background
  scale_fill_okabe_ito()   # <-- colorblind friendly for N Record
```

## Covariates 


```{r filter_collinear}
keep = filter_collinear(present, method = "cor_caret", cutoff = 0.65)
```

```{r add_depth_and_month}
keep = c("depth", "month", keep)
```

Earlier we saved our model input, with background points in obsbkg. Let's make a copy and call it model_input
```{r present_model_input}
model_input = obsbkg
```

Add depth to list of extract Brickman variables
```{r}
present = read_brickman(add = c("depth"))
```

We extract the environmental covariates. 
```{r extract}
variables = extract_brickman(present, model_input, form = "wide")
variables
```

We don't need the .id column. 
```{r wrangle_variables}
variables = variables |>
  mutate(class = model_input$class) |>    # the $ extracts a column 
  select(-.id)                            # the minus means "deselect" or "drop"
variables
```

Let's look at a comparison of the covariates distribution between our presence and background points. 
```{r plot_pres_vs_bg, warning = FALSE}
plot_pres_vs_bg(variables |> select(-month), "class")
```
```{r write_cfg}
cfg = list(
  version = "v3",
  scientificname = SPECIES,
  background = "average of observations per month",
  bias = "random",
  thinning = "false",
  metrics = sdm_metric_set(),
  keep_vars =  keep)


write_configuration(cfg)          
```

And finally, write everything we computed to the disk. 
```{r write_model_input}
write_model_input(variables, scientificname = SPECIES, version = "v3")
```

```{r read model input}
# model_input = read_model_input(scientificname = SPECIES)
# example read model input
```

# Load data - choose a month and sampling approach

```{r load_data}
model_input = read_model_input(scientificname = SPECIES, 
                               version = "v3",
                               log_me = c("depth", "Xbtm")) |>
  dplyr::mutate(month = month_as_number(.data$month)) |>
  select(all_of(c("class", cfg$keep_vars)))
```

# Splitting the data into groups

## Initial split into training and testing groups

```{r initial_split}
model_input_split = spatial_initial_split(model_input, 
                        prop = 1 / 5,     # 20% for testing
                        strategy = spatial_block_cv) # see ?spatial_block_cv
model_input_split
```

```{r initial_split_plot}
autoplot(model_input_split)
```

## Split the training group into regional folds


```{r cv_training}
tr_data = training(model_input_split)
cv_tr_data <- spatial_block_cv(tr_data,
  v = 5,     
  cellsize = grid_cellsize(model_input),
  offset = grid_offset(model_input) + 0.00001
)
autoplot(cv_tr_data)
```


# Build a recipe

```{r recipe}
one_row_of_training_data = dplyr::slice(tr_data,1)
rec = recipe(one_row_of_training_data, formula = class ~ .)
rec
```

```{r recipe_summary}
summary(rec)
```

# Create a workflow


```{r make_workflow}
wflow = workflow_set(
  
  preproc = list(default = rec), # not much happening in our preprocessor
  
  models = list(                 # but we have 4 models to add
    
      # very simple - nothing to tune
      glm = logistic_reg(
          mode = "classification") |>
        set_engine("glm"),
      
      # two knobs to tune
      rf = rand_forest(
          mtry = tune(),
          trees = tune(),
          mode = "classification") |>
        set_engine("ranger", 
                   importance = "impurity"),
      
      # so many things to tune!
      btree = boost_tree(
          mtry = tune(), 
          trees = tune(), 
          tree_depth = tune(), 
          learn_rate = tune(), 
          loss_reduction = tune(), 
          stop_iter = tune(),
          mode = "classification") |>
        set_engine("xgboost"),
    
      # just two again
      maxent = maxent(
          feature_classes = tune(),
          regularization_multiplier = tune(),
          mode = "classification") |>
        set_engine("maxnet")
  )
)
wflow
```


## Chosing metrics for measuring how well the models do


```{r metrics}
metrics = cfg$metrics
metrics
```


# Fit the models to the various recipes

### Iterate to find various hyperparameter sets

```{r fit, warning = FALSE}
wflow <- wflow |>
  workflow_map("tune_grid",
    resamples = cv_tr_data, 
    grid = 3,
    metrics = metrics, 
    verbose = TRUE)
```
Here's a quick plot of the outputs, with each metric being showcased.
```{r plot_wflow}
autoplot(wflow)
```

### Choose the best set of hyperparameters for each model

```{r select_best}
model_fits = workflowset_selectomatic(wflow, model_input_split,
                                  filename = "Temora_longicornis-v3-model_fits",
                                  path = data_path("models"))
model_fits
```

## Exploring the collection of model fit results


### A table of metrics

We can quickly pull together summary statistics.

```{r model_fit_metrics}
model_fit_metrics(model_fits)
```

### Confusion matrices and accuracy

```{r model_fit_confmat}
model_fit_confmat(model_fits)
```
### ROC/AUC


```{r model_fit_roc_auc}
model_fit_roc_auc(model_fits)
```

### Variable importance

```{r model_fit_vip}
model_fit_varimp_plot(model_fits)
```

## Exploring a single model fit result

Let's pull out one of the model fits and look at it's parts.

```{r random_forest}
rf = model_fits |>
  filter(wflow_id == "default_rf")
rf
```

#### `splits`

These are the allocations of training and testing elements in the original dataset. We can see it looks like the same distribution as our initial split (above).

```{r rf_splits}
autoplot(rf$splits[[1]])
```

#### `.metrics`


```{r rf_metrics}
rf$.metrics[[1]]
```

```{r}
autoplot(wflow, metric = "tss")
```
 

#### `.predictions`

Here we can see the **testing** data predictions. Ultimately we are interested in `.pred_class` and `class`, but there's a lot of data to explore in here.  

```{r rf_preds}
rf$.predictions[[1]]
```

#### `.workflow`

The workflow is the encapsulation of the model formula form the recipe, the tuned hyperparameters and well as any fitting coefficients for the mode.  When we print it we get a summary statement about the model.

```{r rf_workflow}
rf$.workflow[[1]]
```


### Partial dependence plot

Partial dependence reflects the relative contribution of each variable influence over it's full range of values.  The output is a grid grid of plots showing the relative distribution of the variable (bars) as well as the relative influence of the variable (line). 
```{r pd_plot_gtree}
model_fit_pdp(model_fits, wid = "default_btree", title = "Boosted Tree")
```

```{r pd_plot_rf}
model_fit_pdp(model_fits, wid = "default_rf", title = "Random Forest")
```

```{r pd_plot_maxent}
model_fit_pdp(model_fits, wid = "default_maxent", title = "Max Entropy")
```

```{r pd_plot_glm}
model_fit_pdp(model_fits, wid = "default_glm", title = "Generalized Linear Model")
```

# Prediction

```{r load_workflow}
model_fits = read_model_fit(filename = "Temora_longicornis-v3-model_fits")
model_fits
```

# Make a prediction

First we shall make a "nowcast" which is just a prediction of the current environmental conditions. 

## Nowcast

First make the prediction using `predict_stars()`.  The function yields a `stars` array object that has one attribute (variable) for each row in the `model_fits` table. We simply provide the covariates (predictor variables) and the function takes care of the rest.

```{r nowcast}
db = brickman_database()
present = read_brickman(db |>  filter(scenario == "PRESENT",  interval == "mon"),
                       add = c("depth", "month")) |>
          select(all_of(cfg$keep_vars))
nowcast = predict_stars(model_fits, present)
nowcast
```
Now we can plot what is often called a "habitat suitability index" (hsi) map. We can 

```{r plot_nowcast_maxent, warning = FALSE}
plot_prediction(nowcast['default_btree'])
```

We can also plot a presence/absence labeled map, but keep in mind it is just a thresholded version of the above where "presence" means `prediction >= 0.5`. Of course, you can select other values to threshold the habitat suitablility scores.

```{r plot_class_labels, warning = FALSE}
pa_nowcast = threshold_prediction(nowcast)
plot_prediction(pa_nowcast['default_btree'])
```

## Forecast

Now let's try our hand at forecasting - let's try RCP85 in 2075.  First we load those parameters, then run the prediction and plot.

```{r load_2075_RCP85, warning = FALSE}
covars_rcp85_2075 = read_brickman(db |> filter(scenario == "RCP85", 
                                               year == 2075, 
                                               interval == "mon"),
                                  add = c("depth", "month")) |>
  select(all_of(cfg$keep_vars))
```

```{r forecast}
forecast_2075 = predict_stars(model_fits, covars_rcp85_2075)
forecast_2075
```

```{r plot_forecast, warning = FALSE}
plot_prediction(forecast_2075['default_btree'])
```


:::{.callout-note}
Don't forget that [there are other ways](https://github.com/BigelowLab/ColbyForecasting/wiki/Spatial) to plot array based spatial data.
:::

## Save the predictions

It's easy to save the predictions (and read then back with `read_prediction()`).

```{r save_pred}
# make sure the output directory exists
path = make_path("predictions")

write_prediction(nowcast,
                 scientificname = cfg$scientificname,
                 version = cfg$version,
                 year = "CURRENT",
                 scenario = "CURRENT")
write_prediction(forecast_2075,
                 scientificname = cfg$scientificname,
                 version = cfg$version,
                 year = "2075",
                 scenario = "RCP85")
```